Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-04-08 manuscript.qmd
add upsetplot

---
title: "Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer"
authors:
  - name: Roman Ludwig
    orcid: 0000-0001-9434-328X
    email: roman.ludwig@usz.ch
  - name: Yoel Perez Haas
    email: yoel.perezhaas@usz.ch
  - name: Jan Unkelbach
    orcid: 0000-0002-4275-990X
    email: jan.unkelbach@usz.ch
format:
  html:
    code-fold: true
  pdf:
    keep-tex: true
    include-in-header:
    - text: |
        \usepackage{multirow}
        \usepackage{centernot}
jupyter: python3
execute: 
  cache: false
crossref:
  fig-prefix: "figure"
  tbl-prefix: "table"
  eq-prefix: "equation"
  sec-prefix: "section"
---

@@

- head and neck cancer spreads lymphatically
- may sometimes spread to contralateral side
- spreads more frequently and to larger extent contralaterally when tumor extends the mid-sagittal line
- we describe a model based on previously published hidden Markov model
- we extend it to cover the contralateral side, too
- naive approach: make two independent models for each side, but that is not supported by the data
- ipsi- and contralateral side are correlated via time of diagnosis, which is correlated with T-category
- tumor extension over mid-sagittal line is modelled as random variable
- spread probabilities from tumor to contralateral LNLs in case of midline extesnion are linear combinations of these probabilities in case of ipsilateral spread and contralateral spread when tumor is clearly lateralized

@@

@@

The predicted prevalence of midline extension for early T-category tumors is `{python} f"{early_midext_prob:.2%}"`. For a late T-category tumor, it is `{python} f"{late_midext_prob:.2%}"`.

In the data, the prevalence of midline extension for early T-category tumors is `{python} f"{early_midext_data:.2%}"`. For late T-category tumors it is `{python} f"{late_midext_data:.2%}"`.

>>>>

---
title: "Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer"
authors:
  - name: Roman Ludwig
    orcid: 0000-0001-9434-328X
    email: roman.ludwig@usz.ch
  - name: Yoel Perez Haas
    email: yoel.perezhaas@usz.ch
  - name: Jan Unkelbach
    orcid: 0000-0002-4275-990X
    email: jan.unkelbach@usz.ch
bibliography: references.bib
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
  pdf:
    keep-tex: true
    include-in-header:
    - text: |
        \usepackage{multirow}
        \usepackage{centernot}
jupyter: python3
execute: 
  cache: true
crossref:
  fig-prefix: "figure"
  tbl-prefix: "table"
  eq-prefix: "equation"
  sec-prefix: "section"
code-links:
  - text: GitHub Repo
    href: https://github.com/rmnldwg/bilateral-paper
    icon: github
resources: 
  - data/reduced.csv
meca-bundle: true
---

@@

- blabla
- head and neck cancer spreads through the lymphatic network
- may sometimes spread to contralateral side
- spreads more frequently and to larger extent contralaterally when tumor extends the mid-sagittal line
- we describe a model based on previously published hidden Markov model
- we extend it to cover the contralateral side, too
- naive approach: make two independent models for each side, but that is not supported by the data
- ipsi- and contralateral side are correlated via time of diagnosis, which is correlated with T-category
- tumor extension over mid-sagittal line is modelled as random variable
- spread probabilities from tumor to contralateral LNLs in case of midline extesnion are linear combinations of these probabilities in case of ipsilateral spread and contralateral spread when tumor is clearly lateralized

@@

# Detailed Data on Lymphatic Involvement

To train our models, we used detailed tabular data specifying the lymphatic involvement of individual patients on a per LNL basis.

```{python}
#| echo: false
#| label: fig-data-upset
#| fig-cap: Upset plot of the ipsilateral lymphatic involvement patterns [@lex_upset_2014]. Here, the horizontal bars depict the overall prevalence of LNL involvemen. I.e., LNL II is the only level that is involved in more than 500 patients. The vertical bar plots depict joint involvements. E.g., the fourth bar from the left depicts how often the LNLs II, III, and IV are jointly involved while level I and V are not. This is indicated by the light and dark shaded dots in the matrix underneath.
#| warning: false
lnls = ["I", "II", "III", "IV", "V"]

indicator_data = upsetplot.from_indicators(lnls, data=data["max_llh", "ipsi"])
upset = upsetplot.UpSet(
  indicator_data,
  sort_by="cardinality",
  sort_categories_by="-input",
  min_subset_size=10,
  facecolor=COLORS["blue"],
)
upset.style_subsets(absent=lnls, facecolor=COLORS["green"], label="N0 (all healthy)")
axes = upset.plot()
axes["totals"].set_xlabel("Prevalence")
axes["intersections"].set_ylabel("Scenario size")
plt.show()
```

@@

The predicted prevalence of midline extension for early T-category tumors is <!-- insert computational code later -->. For a late T-category tumor, it is <!-- insert computational code later -->.

In the data, the prevalence of midline extension for early T-category tumors is <!-- insert computational code later -->. For late T-category tumors it is <!-- insert computational code later -->.
